Disentangling linkages between satellite derived forest structure and productivity essential biodiversity variables.

### 2023-06-05 scripts/4_vegan.qmd
full draft to tx & mea

bec_tab <- keys$bec %>%
  left_join(bec_clim, by = c("zone" = "bec")) %>%
  select(-order) %>%
  arrange(group, zone) %>%
  relocate(group)

>>>>

bec_tab <- keys$bec %>%
  left_join(bec_clim, by = c("zone" = "bec")) %>%
  select(-order) %>%
  arrange(group, zone) %>%
  relocate(group) %>%
  mutate(across(ppt:tmin, \(x) round(x, digits = 1)))

### 2023-06-05 scripts/manuscript/manuscript.qmd
full draft to tx & mea

abstract: In today's changing climate and in light of the biodiversity crises, it is integral to be able to monitor the change in biodiversity at large scales. Key to this is the development of variables capable of monitoring biodiversity trends. One framework developed to tackle this problem is the Essential Biodviersity Variables (EBV), which is designed to be analogous to the Essential Climate variables, and capable of monitoring biodiversity globally while also being complementary to one another, within EBV classes. The EBV framework contains six classes, some of which are extremely well suited to be monitoring from satellite remote sensing. Among these with the potential to be satellite derived, the ecosystem structure and function classes stand out. In this paper, we seek to assess the complementarity of forest structure, as imputed across the entirety of British Columbia, Canada, with the Dynamic Habitat Indices, a yearly summary of productivity indices. Both datasets have been previously linked with biodiversity metrics across a range of scales. Using redundancy analysis, we find that forest structural attributes and the DHIs are decoupled from one another, with the forest structure datasets explaining 14% of the variation in the DHIs. Further, we explore how the proportion of variance explained in the DHIs from primary structural attributres such as canopy height and canopy cover and modelled structural attributes such as aboveground biomass varies. We find there is generally large overlap between the two forest structure datasets. Overall, we find that forest structure as an ecosystem structure EBV and the DHIs as an ecosystem function EBV are well suited to be complementary EBVs across large environmental gradients.

@@

With biodiversity being in decline, and facing extinction rates above
the background extinction rate [@thomas2004; @urban2015], as well as the
homogenization of communities at various scales [@mcgill2015] it is
integral to be able to monitor how biodiversity is changing across the
globe. In response, the global biodiversity community is making efforts
to assess and halt the degradation of biodiversity. The Group for Earth
Observation Biodiversity Observation Network has developed the Essential
Biodiversity Variables [EBVs, @pereira2013], designed as an analog to
the Essential Climate Variables framework [@bojinski2014]. EBVs are
designed to be global in scope, relevant to biodiversity information,
feasible to use, and complementary to one another [@skidmore2021]. While
it can be incredibly difficult, time consuming, and expensive to collect
data on biodiversity across wide swaths of land and varying ecosystems,
EBVs, which can be correlated to sampled biodiversity information, allow
for the monitoring and assessment of protected area effectiveness and
ecosystem health at large spatial scales [@hansen2021]. There are six
EBV classes, each of which correspond to a different facet of
biodiversity including species populations, species traits, community
composition, ecosystem structure, ecosystem function, and genetic
composition [@pereira2013].

@@

The two landscape-level EBVs (ecosystem structure and function) are well
suited to be examined at large spatial scales using coarser spatial
measurements, such as those taken from satellites by the Moderate
Resolution Imaging Spectroradiometer [MODIS; @zhang2003], the Landsat
imaging systems [@fisher2006], or Sentinel-2 [@helfenstein2022;
@darvishzadeh2019] programs. These mid-resolution satellites can monitor
processes at broader extents but the coarse spatial resolution removes
the ability to relate these traits to individual organisms. As a result,
satellite remote sensing data has been shown to be arguably the most
effective at monitoring ecosystem based EBVs focused on structure and
function. These EBVs classes can be monitored at regional to global
extents through the use of optical imagery [@cohen2004], as well as
active sensors such as lidar (light detection and ranging) and radar
[@guo2017; @lefsky2002; @lang2021; @neuenschwander2019; @coops2016].

@@

Forest structural diversity has been linked to biodiversity at various
scales [@guo2017; @bergen2009; @gao2014]. Structural attributes range in
complexity from simple (canopy cover; canopy height), to more complex
(vertical and horizontal structural complexity) to modelled (aboveground
biomass; basal area), all of which can be assessed using lidar data
[@coops2021]. A suite of these lidar-derived attributes have been used
as local indicators of biodiversity, including simple metrics such as
canopy cover and canopy height as well as derived metrics including
vertical profiles, aboveground biomass [@lefsky1999; @guo2017;
@coops2016]. Other second order derived metrics such as canopy texture,
height class distribution, edges, and patch metrics have also been used
to examine habitat and biodiversity at landscape scales [@bergen2009].
Advances in satellite remote sensing processing have allowed 3D forest
structure data to be imputed across wide spatial scales [@matasci2018;
@coops2021] using data fusion approaches involving collected lidar data
and optical/radar data.

@@

@berry2007 first explored this idea by proposing the linkage of
intra-annual summaries of MODIS derived GPP to dispersive bird species.
This idea was further refined into the Dynamic Habitat Indices [DHIs;
@coops2008], which have now been shown to be well suited to assess the
three aforementioned hypotheses at global scales [@radeloff2019]. The
cumulative DHI calculates the total amount of energy available in a
given pixel over the course of a year. Cumulative DHI is strongly linked
to the available energy hypothesis, which suggests that with greater
available energy species richness will increase [@wright1983]. The
minimum DHI, which calculates the lowest productivity over the course of
a year can be matched to the environmental stress hypothesis, which
proposes that higher levels of minimum available energy will lead to
higher species richness [@currie2004]. Finally, the variation DHI, which
calculates the coefficient of variance in a vegetation index through the
course of a year, corresponds to the environmental stability hypothesis
which states that lower energy variation throughout a year will lead to
increased species richness [@williams2008].

@@

Biodiversity monitoring programs often require a range of information in
order to accurately assess changes in ecosystem integrity
[@lindenmayer2010]. Choosing datasets that are most closely related to
the phenomenon of interest in a given application allows for direct
connections to monitoring results and management actions [@pressey2021].
With the advent of large-extent monitoring methods like satellite remote
sensing, and a proliferation of potential EBVs datasets, it becomes
important to assess the interrelationships between these datasets, and
assess their complementary of the information to reduce duplication of
efforts [@pereira2013; @skidmore2021]. When strong relationships are
present between EBV classes, it becomes possible to assess the
ecological relationships between potential EBVs. On the other hand, when
datasets do not appear related, they may be well suited to be used in
monitoring programs together, as complementary EBVs.

@@

The overall goal of this paper is to assess patterns of forest ecosystem
structure and function and their complementarity across a wide range of
ecosystems encompassing significant environmental gradients. To do so,
we synthesize data from moderate-scale remote-sensing derived metrics of
forest structure, represented as both simple ALS-extracted metrics of
canopy height, cover and vertical complexity, as well as modelled forest
structure attributes including volume, and aboveground biomass, with a
well-established remote sensing derived index on ecosystem function. Our
first question is to examine how ecosystem structure and function
complement one another across a large environmental gradient and then
compare the simple and modelled representations of forest structure to
different levels of ecosystem function. This question is important as it
provides insights to the EBV community around complementarity of remote
sensing metrics when describing the structure and function of ecosystems
and proposes a method to examine potential overlap when generating
remote sensing EBVs.

Our second question examines the independent and shared relationships of
ecosystem structure height and cover, with modelled forest structure, on
ecosystem function. This provides insight into the choice of remote
sensing attributes to use when developing EBVs within a single EBV
class. Remote sensing datasets can comprise relatively unprocessed
observations, in this case ALS measures of height and cover which are
derived from the raw 3D point cloud vs modelled attributes, such as
biomass and volume, which involve the use the statistical relationships
with field data to transform the observations into more refined data
products. Assessing which of these two (or combination of the two)
approaches has stronger or weaker correlations with estimates of
function provides insights into the choice of data used to build EBVs.
Lastly, we examine how the primary and modelled structure attributes
partition the variance of the DHIs within key biomes and forest types
across a large environmental range, examining to what extent ecosystem
and forest types impacts these relationships and thus providing insight
into the applicability of these results globally.

@@

We used a suite of forest structural attributes (canopy height, canopy
cover, overstory cover, basal area, aboveground biomass, gross stem
volume, mean elevation, elevation standard deviation, and structural
complexity \[coefficient of variation in elevation returns\]). This
dataset was created for the year 2015 at a 30 m spatial resolution
according to @matasci2018. In brief, the method used a set of lidar
collections and field plots across Canada, and imputed the remaining
pixels using a random forest k-Nearest Neighbour approach on
Landsat-derived surface reflectance and auxiliary data such as
topography. Detailed information on the creation of this dataset can be
found in @matasci2018.

@@

We use an established set of indices of annual productivity shown to be
related to global biodvierstiy trends: the Dynamic Habitat Indices
[@radeloff2019]. The DHIs are a set of satellite remote sensing derived
productivity variables that summarize the cumulative amount of available
energy, the minimum available energy, and the variation in available
energy throughout a given year [@berry2007; @radeloff2019]. The DHIs
have previously been produced at a global extent using MODIS imagery ,
and have been used to assess alpha [@radeloff2019] and beta[@andrew2012]
diversity, species abundances [@razenkova2023], and construct novel
ecoregionalizations [@coops2009]; @andrew2013\] . Recent studies have
began to examine how these indices can be constructed at a finer spatial
resolution by using multi-annual Landsat imagery to generate a single
synthetic year of monthly observations [@razenkova2022].

The DHIs were calculated according to [@razenkova2022; Razenkova et al.,
In Press] for all of terrestrial British Columbia. In brief, Google
Earth Engine [@gorelick2017] was used to obtain all valid Landsat pixels
for a given study area, filtering out pixels containing shadows, slouds,
and cloud shadows within each image [@zhu2012], then calculated the NDVI
for each pixel in each image. They then calculated the median NDVI value
for each month across the ten year time span (2011-2020) to generate a
synthetic year of monthly data. The sum, minimum, and coefficient of
variation across this synthetic year of NDVI values is then calculated.
More detailed information can be found in Razenkova et al. (In Press).

## Sampling

We conducted model-based sampling across the sixteen ecosystems found
within British Columbia (@fig-study-map). Samples were randomly selected
within each BEC zone, in undisturbed pixels. Each sampled pixel had to
have a forested land cover class (coniferous, deciduous, mixed-wood, or
wetland-treed), and be surrounded by the same land cover class. The land
cover mask was generated following [@hermosilla2022] using a
best-available-pixel composite, and an inverse-distance weighted random
forest approach across Canada. Additionally, each pixel had to have a
coefficient of variation less than 0.5 in surrounding pixels in the two
simplest forest structural attributes, canopy height and canopy cover. A
maximum of 3000 samples were sampled in each BEC zone with a 1 km
minimum sampling distance to reduce the effects of spatial
autocorrelation. All variables were natural-log transformed and
standardized. Variables containing zeros were natural-log plus one
transformed. Sampling was conducted in R [@R-base] version 4.2.2 using
the **sgsR** package [@R-sgsR]. Focal analyses for the land cover
classes and coefficient of variation were calculated in Python version
3.9.

# Sampling 2

We implement two sampling schemes -- one for ecosystems, and one for
forest types -- across the terrestrial ecosystems of British Columbia
(@fig-study-map). Both have samples randomly selected. For both sampling
schemes, each individual sample was taken from a forested pixel which
was surrounded by the same land cover class. The land cover class map
was generated for the year 2015 following [@hermosilla2022] using a
best-available-pixel composite, and an inverse-distance weighted random
forest approach across Canada. Additionally, each pixel had to have a
coefficient of variation less than 0.5 in surrounding pixels in the two
simplest forest structural attributes, canopy height and canopy cover. A
minimum sampling distance of 1 km was implemented to reduce the effects
of spatial autocorrelation, and samples that had been disturbed in the
last 30 years were discarded by using a disturbance mask generated for
the forested ecosystems of Canada by @hermosilla2016.

The ecosystem sampling scheme was implemented by randomly sampling
pixels matching the above criteria within each BEC zone of British
Columbia, up to a maximum of 3000 pixels, regardless of land cover
class. The forest type sampling scheme was similar in that each BEC zone
was sampled, however, it implemented stratified sampling on each forest
type, up to a maximum of 500 pixels per BEC zone and forest type.

@@

Redundancy analysis (RDA) and variation partitioning were used to relate
the primary and modelled forest structure attributes to ecosystem
function across a broad environment range. Redundancy analysis functions
similarly to a multiple linear regression, except it is capable of
predicting multiple response variables. It accomplishes this by first
running a multiple linear regression of each predictor variable on each
response variable, then running a principle component analysis on the
residuals from each multiple linear regression. This reduces the
dimensionality of the output, and allows the relationship strength to be
assessed by calculating the loadings of both predictor and response
variables on the RDA axes. Partial redundancy analysis function
similarly, except also considers co-variates [@legendre2012]. Redundancy
analysis has widely been used in community ecology where environmental
variables of interest are compared to species composition
[@blanchet2014; @kleyer2012] and has similarities to partial least
squares regression -- namely, their multivariate approach, usage of
dimensionality reduction, and linearity assumptions -- which is commonly
used in remote sensing literature [@roelofsen2014; @burnett2021].

Following the RDA, we employ ANOVAs to determine which axes are
significant, and calculate the proportion of variance attributable to
each axis using the eigenvalues generated from the RDA. We calculate
axis loadings for both predictor and response variables by calculating
the correlation between the variables and the RDA axes. Axis loadings
represent the relationship between a given variable and the RDA axis. We
only consider and display significant axes. To visualize the RDA for
both predictor and response variables, we display the results as path
diagrams, with loadings from each predictor to the RDA axis to the
response variables. The variance explained by each axis is also
displayed in the RDA box (@fig-rda-var). All RDA calculations were done
in R [@R-base] version 4.2.2 using the **vegan** package [@R-vegan].

Variation partitioning is an extension of partial RDA which can assess
the overlap between the explanatory power of two datasets by utilizing
multiple partial RDAs and exchanging which datasets are considered the
predictor, and which is considered the co-variate [@legendre2012].
Variation partitioning is traditionally displayed using a Venn diagram,
in which the percentage of variance explained by each dataset is in a
circle, and the overlap between circles represents the overlap in
variance explained. All variation partition analyses were done in R
[@R-base] version 4.2.2 using the **vegan** package [@R-vegan].

RDA and variation partitioning analyses were conducted for all samples
in the ecosystem sampling scheme, as well as individually across each
ecosystem and forest type. The results were aggregated to BEC zone
groups (see @sec-study-area).

@@

```{r}
#| label: fig-rda-var
#| fig-cap: A) Axis loadings from redundancy analysis of primary and modelled forest structural attributes on the dynamic habitat indices. B) Results from variation partitioning of primary and modelled forest structural attributes on the DHIs. Both visualized analyses are across all collected samples. See supplementary information for results from each BEC zone and forest type.

@@

To examine the relationship between ecosystem structure and function
across a large environmental gradient, we present the results of a
redundancy analysis of forest structural attributes on the dynamic
habitat indices across the entire sampled dataset in @fig-rda-var A.
While there are three RDA axes associated with the full dataset, the
third axis explains 0.05% of the variance in the DHIs, and as a result
it is not shown. The first axis is strongly represents all the DHIs
(loadings \> 0.85 for all DHIs), and has the strongest loadings from
canopy cover, basal area, aboveground biomass, and gross stem volume.
The other input attributes (canopy height, structural complexity) have
smaller loadings. The second axis primarily represents the seasonality
(Minimum and Variation DHI) of the DHIs, with a very small loading on
the Cumulative DHI, and has the largest loadings in canopy cover and
complexity, with the rest of the forest structural attributes having
negligible loadings (@fig-rda-var A).

The results from the variation partitioning analysis (@fig-rda-var B)
show that the majority of the variance explained by the input datasets
is shared across both primary and modelled attributes. In total, only
13.4% of the variation in the DHIs is explained by the structural
information, with 9.4% of this being from overlap between the datasets.
The primary and modelled attributes explain 2.7 and 1.3% of the
variation on their own, respectively (@fig-rda-var B).

@@

Further, we wanted to examine the strength of the relationship between
the DHIs and primary structural attributes vs modelled structural
attributes across the forest types and ecosystems of the province.
@fig-radar A and B show the axis loadings for structural attributes and
the DHIs, respectively. Across the BEC zone groups, the loadings are
generally similar in the structural attributes. Southern ecosystems have
low or non-significant structural complexity loadings, while the forest
types have structural complexity as their largest loading. The loadings
for the DHIs in the first axis are generally larger than the loadings
for the structural attributes, with the cumulative DHI often being the
strongest loading overall. In alpine BEC zones, the minimum DHI loading
is smaller than the variation and cumulative DHIs. In contrast to the
first RDA axis, the secondary RDA axis is primarily driven by variation
in the minimum DHI, with medium loadings in the variation DHI, and small
loadings in the cumulative DHI (@fig-radar A & B). @fig-radar C shows
boxplots of individual BEC zone loadings for structural attributes and
the DHIs. Across the BEC zones, canopy cover is generally the strongest
predictor loading of the first RDA axis, and structural complexity is
generally the weakest predictor loading. The exception is in the forest
types, where structural complexity is often the largest driver of the
first axis. When significant, the second RDA axis always has small
structural attribute loadings.

```{r}
#| label: fig-fracts-bar
#| fig-cap: Stacked bar plot proportion of DHI variation explained by extracted, modelled, and the overlap between primary and modelled structural attributes.

@@

@fig-fracts-bar shows stacked bar plots of the proportion of variation
explained in the DHIs by primary, modelled, and the overlap between
primary and modelled attributes. The forest types sampling scheme
generally results in higher amounts of variance explained, which is
especially prevalent in broadleaf and mixed wood forests. The overlap
between modelled and primary attributes is generally higher than each
dataset on their own, with the exception being mixed wood and broadleaf
forests and the Coastal Western Hemlock and Engellmann Spruce - -
Subalpine Fir BEC zone zones. The Interior Mountain-heather Alpine and
Montane Spruce ecosystems have larger amounts of DHI variance explained
than the rest of their ecosystem groups. Notably, the variation
explained by the primary and modelled attributes is commonly presented
through a single RDA axis, which generally corresponds to overall
productivity through the year (@fig-radar). Generally, the DHIs are
decoupled from the structural attributes, as shown by the overall
variance explained being under 30%, regardless of whether the attribute
is directly extracted from the point cloud or is a modelled variable.

@@

knitr::include_graphics(here::here("outputs", "rgb_norm_edited.png"))
```

>>>>

abstract: Under a changing climate and in light of the biodiversity crises, monitoring change in biodiversity at broad scales has become ever more imperative. Key to regular biodiversity monitoring over time and space is the definition and selection of variables capable of accurately monitoring biodiversity trends. One framework is the Essential Biodiversity Variables (EBV), which are designed to be analogous to the Essential Climate variables, and capable of monitoring biodiversity globally while also being complementary to one another, within EBV classes. Five of six EBV classes have been shown to be well suited to be monitored using satellite remote sensing, with particular focus warranted on the ecosystem structure and function classes. In this paper, we seek to address the complementarity of variables describing forest structure, as imputed across the entirety of British Columbia, Canada, with variables representing ecosystem functioning, namely the Dynamic Habitat Indices, a yearly summary of productivity indices. Both sets of variables have previously been linked with biodiversity metrics across a range of scales. Using redundancy analysis, we find that forest structural attributes and the DHIs are essentially decoupled from each other, with the forest structure variables explaining 14% of the variation in the DHIs. Further, we explore how the proportion of variance explained in the DHIs varies with either primary structure variables, such as canopy height and cover, vs more complex modelled forest structural attributes, such as aboveground biomass and volume. Results suggest there is large overlap between the two types of forest structure variables suggesting little added benefit of derived attributes beyond height and cover, in this context. Overall, we find that forest structure as an ecosystem structure EBV and the DHIs as an ecosystem function EBV represent markedly different processes as observed by remote sensing and they are highly complementary across large environmental gradients, suggesting that in an earth-observation based monitoring program both are needed.

@@

With biodiversity being in decline, and facing extinction rates above
the background extinction rate [@thomas2004; @urban2015], as well as the
homogenization of communities at various scales [@mcgill2015] it is
integral to be able to monitor how biodiversity is changing across the
globe. In response, the global biodiversity community is making efforts
to assess and halt the degradation of biodiversity. The Group for Earth
Observation Biodiversity Observation Network has developed the Essential
Biodiversity Variables [EBVs, @pereira2013], designed as an analog to
the Essential Climate Variables framework [@bojinski2014]. EBVs are
designed to be global in scope, relevant to biodiversity information,
feasible to use, and complementary to one another [@skidmore2021]. While
it can be difficult, time consuming, and expensive to collect data on
biodiversity across wide swaths of land and varying ecosystems, EBVs,
which can be correlated to sampled biodiversity information, allow for
the monitoring and assessment of protected area effectiveness and
ecosystem health at large spatial scales [@hansen2021]. There are six
EBV classes, each of which correspond to a different facet of
biodiversity, including species populations, species traits, community
composition, ecosystem structure, ecosystem function, and genetic
composition [@pereira2013].

@@

The two landscape-level EBVs (ecosystem structure and function) are well
suited to be examined at large spatial scales using coarser spatial
measurements, such as those taken from satellites by the Moderate
Resolution Imaging Spectroradiometer [MODIS, @zhang2003], the Landsat
imaging systems [@fisher2006], or Sentinel-2 [@helfenstein2022;
@darvishzadeh2019] programs. These mid-resolution satellites can monitor
processes at broader extents but the coarse spatial resolution removes
the ability to relate these traits to individual organisms. As a result,
satellite remote sensing data has been shown to be arguably the most
effective at monitoring ecosystem based EBVs focused on structure and
function. These EBVs classes can be monitored at regional to global
extents through the use of optical imagery [@cohen2004], as well as
active sensors such as lidar (light detection and ranging) and radar
[@guo2017; @lefsky2002; @lang2021; @neuenschwander2019; @coops2016].

@@

Forest structural diversity has been linked to biodiversity at various
scales [@guo2017; @bergen2009; @gao2014]. Structural attributes range in
complexity from simple (canopy cover; canopy height), to more complex
(vertical and horizontal structural complexity) to modelled (aboveground
biomass; basal area), all of which can be assessed using lidar data
[@coops2021]. A suite of these lidar-derived attributes have been used
as local indicators of biodiversity, including simple metrics such as
canopy cover and canopy height as well as derived metrics including
vertical profiles, aboveground biomass, or lidar derived indices
[@lefsky1999; @guo2017; @coops2016]. Other second order derived metrics
such as canopy texture, height class distribution, edges, and patch
metrics have also been used to examine habitat and biodiversity at
landscape scales [@bergen2009]. Advances in satellite remote sensing
processing have now allowed 3D forest structure data to be imputed
across wide spatial scales [@matasci2018; @coops2021] using data fusion
approaches involving collected lidar data and optical/radar data.

@@

@berry2007 first explored this idea by proposing the linkage of
intra-annual summaries of MODIS derived GPP to dispersive bird species.
This idea was further refined into the Dynamic Habitat Indices [DHIs,
@coops2008], which have now been shown to be well suited to assess the
three aforementioned hypotheses at global scales [@radeloff2019]. The
cumulative DHI calculates the total amount of energy available in a
given pixel over the course of a year. Cumulative DHI is strongly linked
to the available energy hypothesis, which suggests that with greater
available energy species richness will increase [@wright1983]. The
minimum DHI, which calculates the lowest productivity over the course of
a year can be matched to the environmental stress hypothesis, which
proposes that higher levels of minimum available energy will lead to
higher species richness [@currie2004]. Finally, the variation DHI, which
calculates the coefficient of variance in a vegetation index through the
course of a year, corresponds to the environmental stability hypothesis
which states that lower energy variation throughout a year will lead to
increased species richness [@williams2008].

@@

Biodiversity monitoring programs often require a range of information in
order to accurately assess changes in ecological integrity
[@lindenmayer2010]. Choosing datasets that are most closely related to
the phenomenon of interest in a given application allows for direct
connections to monitoring results and management actions [@pressey2021].
With the advent of large-extent monitoring methods like satellite remote
sensing, and a proliferation of potential EBVs datasets, it becomes
important to assess the interrelationships between these datasets, and
assess their complementary of the information to reduce the duplication
of efforts [@pereira2013; @skidmore2021]. When strong relationships are
present between EBVs, it is possible to assess the ecological
relationships between potential EBVs. On the other hand, when datasets
do not appear related, they may be well suited to be used in monitoring
programs together, as complementary EBVs.

@@

The overall goal of this paper is to assess patterns of forest ecosystem
structure and function, as observed using remote sensing technologies,
and their complementarity across a wide range of ecosystems encompassing
significant environmental gradients. To do so, we synthesize data from
moderate-scale remote-sensing derived metrics of forest structure,
represented as both simple ALS-extracted metrics of canopy height, cover
and vertical complexity, as well as modelled forest structure attributes
including volume, and aboveground biomass, with a well-established
remote sensing derived index on ecosystem function. First, we examine
how ecosystem structure and function complement one another across a
large environmental gradient and then compare the simple and modelled
representations of forest structure to different levels of ecosystem
function. This question is important as it provides insights to the EBV
community around complementarity of remote sensing metrics when
describing the structure and function of ecosystems and proposes a
method to examine potential overlap when generating remote sensing EBVs.

Our second question examines the independent and shared relationships of
ecosystem structure height and cover, with modelled forest structure, on
ecosystem function. This provides insight into the choice of remote
sensing attributes to use when developing EBVs within a single EBV
class. Remote sensing datasets can comprise relatively unprocessed
observations, in this case ALS measures of height and cover which are
derived from the raw 3D point cloud vs modelled attributes, such as
biomass and volume, which involve the use the statistical relationships
with field data to transform the observations into more refined data
products. Assessing which of these two (or combination of the two)
approaches has stronger or weaker relationships with estimates of
ecosystem function provides insights into the choice of data used to
build EBVs. Lastly, we examine how the primary and modelled structure
variables partition the variance of the DHIs within key biomes and
forest types across a large environmental range, examining to what
extent ecosystem and forest types impacts these relationships and thus
providing insight into the applicability of these results globally.

@@

We used a suite of forest structure variables (canopy height, canopy
cover, structural complexity \[coefficient of variation of height
returns\], basal area, aboveground biomass, and gross stem volume). This
dataset was created for the year 2015 at a 30 m spatial resolution
according to @matasci2018. In brief, the method used a set of lidar
collections and field plots across Canada, and imputed the remaining
pixels using a random forest k-Nearest Neighbour approach on
Landsat-derived surface reflectance and auxiliary data such as
topography. Detailed information on the creation of this dataset can be
found in @matasci2018.

@@

We use an established set of indices of annual productivity shown to be
related to global biodiversity trends: the Dynamic Habitat Indices
[@radeloff2019]. The DHIs are a set of satellite remote sensing derived
productivity variables that summarize the cumulative amount of available
energy, the minimum available energy, and the variation in available
energy throughout a given year [@berry2007; @radeloff2019]. The DHIs
have previously been produced at a global extent using MODIS imagery,
and have previously been used to assess alpha [@radeloff2019] and beta
[@andrew2012] diversity, species abundances [@razenkova2023], and
construct novel ecoregionalizations [@coops2009; @andrew2013] . Recent
studies have began to examine how these indices can be constructed at a
finer spatial resolution by using multi-annual Landsat imagery to
generate a single synthetic year of monthly observations
[@razenkova2022].

The DHIs were calculated according to [@razenkova2022, Razenkova et al.,
In Press] for all of terrestrial British Columbia. In brief, Google
Earth Engine [@gorelick2017] was used to obtain all valid Landsat pixels
for a given study area, filtering out pixels containing shadows, clouds,
and cloud shadows within each image using the fmask algorithm
[@zhu2012], then calculated the NDVI for each pixel in each image. They
then calculated the median NDVI value for each month across a ten year
time span (2011-2020) to generate a synthetic year of monthly data. The
sum, minimum, and coefficient of variation across this synthetic year of
NDVI values was then calculated. More detailed information can be found
in Razenkova et al. (In Press).

A summary of remote sensing derived datasets can be found in Table 2.

# Sampling

We implement two sampling schemes -- one for ecosystems, and one for
forest types -- across the terrestrial ecosystems of British Columbia
(@fig-study-map). For both sampling schemes, each individual sample was
taken from a forested pixel which was surrounded by the same land cover
class. The land cover class map was generated for the year 2015
following [@hermosilla2022] using a best-available-pixel composite, and
an inverse-distance weighted random forest classification approach
across Canada. Additionally, each pixel had to have a coefficient of
variation less than 0.5 in surrounding pixels in the two simplest forest
structure variables, canopy height and canopy cover. A minimum sampling
distance of 1 km was implemented to reduce the effects of spatial
autocorrelation, and samples that had been disturbed in the last 30
years were discarded by using a disturbance mask generated for the
forested ecosystems of Canada by @hermosilla2016.

The ecosystem sampling scheme was implemented by randomly sampling
pixels matching the above criteria within each BEC zone of British
Columbia, up to a maximum of 3000 pixels, regardless of forest class.
The forest type sampling scheme was similar in that each BEC zone was
sampled, however, it implemented stratified sampling on each forest
class, up to a maximum of 500 pixels per BEC zone and forest class.

@@

Redundancy analysis (RDA) and variation partitioning were used to relate
the primary and modelled forest structure variables to ecosystem
function across a broad environment range. Redundancy analysis functions
similarly to a multiple linear regression, except it is capable of
predicting multiple response variables. It accomplishes this by first
running a multiple linear regression of each predictor variable on each
response variable, then running a principle component analysis on the
residuals from each multiple linear regression. This reduces the
dimensionality of the output, and allows the relationship strength to be
assessed by calculating the loadings of both predictor and response
variables on the RDA axes. Partial redundancy analysis functions
similarly, except also considers co-variates [@legendre2012]. Redundancy
analysis has widely been used in community ecology where environmental
variables of interest are compared to species composition
[@blanchet2014; @kleyer2012] and has similarities to partial least
squares regression -- namely, their multivariate approach, usage of
dimensionality reduction, and linearity assumptions -- which is commonly
used in remote sensing literature [@roelofsen2014; @burnett2021].

Following the RDA, we employ ANOVAs to determine which axes are
significant, and calculate the proportion of variance attributable to
each axis using the eigenvalues generated from the RDA. We calculate
axis loadings for both forest structure and DHI variables by calculating
the correlation between each variable and the RDA axes. Forest strcuture
variable axis loadings represent the strength of relationship between a
given variable and the RDA axis, while DHI loadings indicate what is
being represented by the RDA axes. We only consider and display
significant axes. To visualize the RDA for both predictor and response
variables, we display the results as path diagrams, with loadings from
each predictor to the RDA axis to the response variables. The variance
explained by each axis is also displayed in the RDA box (@fig-rda-var).

Variation partitioning is an extension of partial RDA which can assess
the overlap between the explanatory power of two datasets by utilizing
multiple partial RDAs and exchanging which datasets are considered the
predictor, and which is considered the co-variate [@legendre2012].
Variation partitioning is traditionally displayed using a Venn diagram,
in which the percentage of variance explained by each dataset is in a
circle, and the overlap between circles represents the overlap in
variance explained.

RDA and variation partitioning analyses were conducted for all samples
in the ecosystem sampling scheme, as well as individually across each
ecosystem and forest type. The results were aggregated to BEC zone
groups (see @sec-study-area). All RDA and variation partitioning
calculations were done in R [@R-base] version 4.2.2 using the **vegan**
package [@R-vegan].

@@

```{r}
#| label: fig-rda-var
#| fig-cap: A) Axis loadings from redundancy analysis of primary and modelled forest structure variables on the dynamic habitat indices. B) Results from variation partitioning of primary and modelled forest structure variables on the DHIs. Both visualized analyses are across all collected samples. See supplementary information for results from each BEC zone and forest type.

@@

To examine the relationship between ecosystem structure and function
across a large environmental gradient, we present the results of a
redundancy analysis of forest structure variables on the dynamic habitat
indices across the entire sampled dataset in @fig-rda-var A. While there
are three RDA axes associated with the full dataset, the third axis
explains 0.05% of the variance in the DHIs, and as a result is not
shown. The first axis strongly represents all the DHIs (loadings \> 0.85
for all DHIs), with the highest loadings from canopy cover, basal area,
aboveground biomass, and gross stem volume. The other input variables
(canopy height, structural complexity) have smaller loadings. The second
axis primarily represents the seasonality (Minimum and Variation) of the
DHIs, with no correlation to Cumulative DHI, and has the highest
loadings from canopy cover and complexity, with the remainder of the
forest structure variables having negligible influences(@fig-rda-var A).

The results from the variation partitioning analysis (@fig-rda-var B)
show that the majority of the variance explained by the input datasets
is shared across both primary and modelled forest structure variables.
Of the 13.4% of the variation in the DHIs being explained by the
structural information, 9.4% is common across both sets of structure
variables. The primary and modelled variables only explain 2.7% and 1.3%
of the variation uniquely (@fig-rda-var B).

@@

Further, we wanted to examine the strength of the relationship between
the DHIs and primary structure variables vs modelled structure variables
across the forest types and ecosystems of the province. @fig-radar A and
B show the axis loadings for structure variables and the DHIs,
respectively. Across the BEC zone groups, the loadings are generally
similar in the structure variables. Variations in the DHIs in the
southern ecosystems, which are typically warmer and drier, have low or
non-significant variance explained by the structure variables.
Conversely, across the forest types (coniferous, broadleaf, mixed wood,
and wetland-treed) structural complexity is often the highest loading.
The loadings for the DHIs in the first axis are generally larger than
the loadings for the structure variables, with the cumulative DHI often
being the strongest loading overall. In alpine BEC zones, the minimum
DHI loading is smaller than the variation and cumulative DHIs. In
contrast to the first RDA axis, the secondary RDA axis is primarily
driven by variation in the minimum DHI, with medium loadings in the
variation DHI, and small loadings in the cumulative DHI (@fig-radar A &
B). @fig-radar C shows boxplots of individual BEC zone loadings for
structure variables and the DHIs. Across the province, canopy cover
generally has the highest loadings, with structural complexity generally
being the weakest.

```{r}
#| label: fig-fracts-bar
#| fig-cap: Stacked bar plot proportion of DHI variation explained by extracted, modelled, and the overlap between primary and modelled structure variables.

@@

@fig-fracts-bar shows stacked bar plots of the proportion of variation
explained in the DHIs by primary, modelled, and the overlap between
primary and modelled variables. Forest types sampled across the
environmental gradient generally show higher amounts of variance
explained, which is especially prevalent in broadleaf and mixed wood
forests. The overlap between modelled and primary structure variables is
generally than the primary and modelled attributes individually, with
the exception being mixed wood and broadleaf forests and colder, wetter
forests. Higher elevation and colder forest stands have larger amounts
of DHI variance explained by the forest structure variables than other
environments in the province. Notably, the variation explained by the
primary and modelled structure variables is commonly presented through a
single RDA axis, which generally corresponds to overall productivity
through the year (@fig-radar). Overall, the DHIs are decoupled from the
structure variables, as shown by the overall variance explained being
under 30%, regardless of whether the attribute is directly extracted
from the point cloud or is a modelled variable.

@@

knitr::include_graphics(here::here("outputs", "rgb_norm_edit.png"))
```

@fig-fcc shows false colour composites of the primary (A & C) and
modelled (B & D) loadings across the ecosystems of British Columbia. In
the first RDA axis there is spatial variation in the primary structure
variables, with interior zones being primarily driven by canopy cover
(green), the coastal regions having strong structural complexity and
canopy height loadings, and the boreal in the northwest having the
strongest loading in canopy height (@fig-fcc A). The modelled structure
variables generally show gray-scale colour, indicating that basal area,
total biomass, and gross stem volume explain similar variation in the
DHIs (@fig-fcc B). All ecosystems with a secondary axis are located away
from the Pacific coast in cooler, drier environments. Canopy cover has
the strongest secondary axis loading in the northern boreal stands,
while the secondary axis in the remaining four zones have the highest
loadings in canopy height (@fig-fcc C). Again, the modeled structure
variables are in grayscale, indicating similar loadings across the three
modelled structure variables (@fig-fcc E).
